Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-24 paper/paper.tex
Section 4.2

%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
The extracted OCaml code is a single recursive traversal. Note the similarity to
a hand-written implementation, and that it has been derived by fusing two
different recursive programs using regular Coq rewrite tactics.

\subsection{Knapsack}\label{sec:knap}
We show now show the formalisation and extraction of \emph{dynamorphisms} for
dynamic programming using the knapsack example from~\cite{HinzeWG15}. In
general, dynamorphisms build a memoisation table which stores intermediate
results alongside the current computation. The algebra used to define a
dynamorphism can access this memoisation table to speed up computation.
First, we define memoisation tables in terms of containers:
\begin{coqcode}
Instance Memo : Cont (A * Sg) Pg := { valid := valid \o pair (snd \o fst) snd }.
Definition Table := LFix Memo.
\end{coqcode}
Given a type of shapes \coq{Sg} and positions \coq{Pg}, the container
\coq{Memo : Cont (A * Sg) Pg}.  The least fixed-point of which is the least
fixed-point of the

given a container \coq{G : Cont Sg Pg}

We define a function to insert elements into the memo tables:
\begin{coqcode}
Definition Cons : A * App G Table ~> App Memo Table := (* *)
\end{coqcode}
And two functions, to inspect the head of a table, and remove an
element:
\begin{coqcode}
Definition headT : Table ~> A := (* *)
Definition tailT : Table ~> App G Table := (* *)
\end{coqcode}
These tables map ``paths'' in the least fixed point of \coq{Memo} to elements
of type \coq{A}. For example, if \coq{G} is a list-generating functor, these
paths will be natural numbers. 
Using these definitions, a dynamorphism is defined as follows:
\begin{coqcode}
Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
  := headT \o hylo (l_in \o Cons \o pair a id) c.
\end{coqcode}
Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
memo table. The definition of the algebra can use this table to lookup
elements, instead of triggering a further recursive call. Elements are inserted
into the memoisation table by the use of \coq{Cons} to the result of applying
the algebra. The algebra for the dynamorphism looks up the previously computed
elements to produce the result, thus saving the corresponding recursive calls:
\begin{coqcode}
Fixpoint memo_knap table wvs :=
  match wvs with | nil => nil | h :: t =>
      match lookupT (Nat.pred (fst h)) table with
      | Some u => (u + snd h)%sint63 :: memo_knapsack table t
      | None => memo_knapsack table t
      end
  end.
Definition knapsack_alg (wvs : list (nat * int))
  (x : App NatF (Table NatF int)) : int :=
  match x with | MkCont sx kx => match sx with
  | inl tt => fun kx => 0%sint63
  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
  end kx end.
Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
\end{coqcode}
Given the recursive coalgebra for \coq{nat}, \coq{out_nat}, knapsack is defined
as
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
While the extracted code has the expected performance characteristics, Coq's
code extraction mechanism is unable to inline several definitions in this case.
We have manually inlined the extracted code for simplicity. The full extracted
code is available in the artefact accompanying this paper, and an interested
reader can check that simple inlining would produce the following function:
% The reader can check
% in our artefact that the extracted code can be trivially inlined to produce the
% following:
\begin{ocamlcode}
let knapsack wvs x = let (y, _) = (let rec f x0 =
      if x0=0 then
        { lFix_out = {shape = Uint63.of_int (0); cont  = fun _ -> f 0 } }
      else let fn := f (x0-1) in { lFix_out = {
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape in y
\end{ocamlcode}
Note how the recursive calls of \ocaml{f} build the memoisation table, and how
this memoisation table is used to compute the intermediate results in
\ocaml{memo_knap}, which is finally discarded to produce the final result.

>>>>

%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
The extracted OCaml code is a single recursive traversal. Note the similarity to
a hand-written implementation, and that it has been derived by fusing two
different recursive programs using regular Coq rewrite tactics.
\newcommand{\Id}{\text{Id}}
\subsection{Knapsack}\label{sec:knap}
Dynamorphisms are hylomorphisms where the algebra has access to the memory table
of the result of all previous recursive calls. To implement this we need to
formalise the idea of a memory table which is in our case simply represented by
the cofree comonad. For a functor $G : \CatC \to \CatC$ the cofree comonad is
defined as the greatest solution to the following equation
\[
  G_{\infty} A \iso A \times G (G_{\infty} A)
\]
In words, these are the (possibly) infinite trees which branch out with shape
$G$. For example, for the identity functor $\Id_{\infty} A$ is the type of
infinite streams. For $G X = 1 + X$, the type $G_{\infty} A$ is isomorphic to
$i \times (1 + G_{\infty} A)$. This is the type representing our memory table
with the operation $\head_{A} : G_{\infty} A$ being the result of the previous
recursive call. Notice that in the case of a recursive $G$-coalgebra this type
only contains finite $G$-trees and so we can view $G_{\infty}A$ as the finite
$G$-trees thus being able to inspect all the previous recursive calls. This has
been expounded thoroughly by Hinze et al.~\cite{HinzeWG15}.

In our framework, we formalise $G$ as a container with shape \coq{Sg} and
position \coq{Pg}, then \coq{Memo} is the least fixed-point of the functor
$F X = A \times GX$:
\begin{coqcode}
Instance Memo : Cont (A * Sg) Pg := { valid := valid \o pair (snd \o fst) snd }.
Definition Table := LFix Memo.
\end{coqcode}
The operations of the cofree comonad \coq{Memo} are the \coq{Cons}, the \coq{headT} and the \coq{tailT}:
\begin{coqcode}
Definition consT : A * App G Table ~> Table := (* *)
Definition headT : Table ~> A := (* *)
Definition tailT : Table ~> App G Table := (* *)
\end{coqcode}
Finally, we define a dynamorphism as a hylomorphism from a type \coq{B} into a
type \coq{Table} post-composed with the \coq{headT} map which outputs the last
result from the memory table:
\begin{coqcode}
Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
  := headT \o hylo (consT \o pair a id) c.
\end{coqcode}
Note the map \coq{App G Table ~> A} corresponding to a map $G G_{\infty}A \to A$ is not a $G$-algebra.
To recover the $G$-algebra we pair it with the identity map and  so we need to
transmogrify it by pairing it with the identity map and post-composing it with
\coq{consT}. This algebra uses this table to lookup
elements, instead of triggering a further recursive call, then elements are inserted
into the memoisation table by the use of \coq{consT} to the result of applying
the algebra. In our example this algebra is called \coq{knapsackA} (see accompanying code).
% \begin{coqcode}
% Fixpoint memo_knap table wvs :=
%   match wvs with | nil => nil | h :: t =>
%       match lookupT (Nat.pred (fst h)) table with
%       | Some u => (u + snd h)%sint63 :: memo_knapsack table t
%       | None => memo_knapsack table t
%       end
%   end.
% Definition knapsack_alg (wvs : list (nat * int))
%   (x : App NatF (Table NatF int)) : int :=
%   match x with | MkCont sx kx => match sx with
%   | inl tt => fun kx => 0%sint63
%   | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
%   end kx end.
% Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
% \end{coqcode}
Finally, given the recursive coalgebra \coq{out_nat} on the natural numbers, we can define knapsack using the recursion scheme for dynamorphisms:
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
The full extracted code coming from this specification
is available in the artefact accompanying this paper, and an interested reader
can check that simple inlining would produce the following function:
% The reader can check
% in our artefact that the extracted code can be trivially inlined to produce the
% following:
\begin{ocamlcode}
let knapsack wvs x = let (y, _) = (let rec f x0 =
      if x0=0 then
        { lFix_out = {shape = Uint63.of_int (0); cont  = fun _ -> f 0 } }
      else let fn := f (x0-1) in { lFix_out = {
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape in y
\end{ocamlcode}
Note that in the extracted code, the recursive calls to \ocaml{f} build the memoisation table, and that
this memoisation table is used to compute the intermediate results in
\ocaml{memo_knap}, which is finally discarded to produce the final result.
